---
title: "OISST Data"
author: "cywhale"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    highlight: tango
    self-contained: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy = FALSE)
options(htmltools.dir.version = FALSE)
options(useFancyQuotes = FALSE)
```

Marine Heatwaves (MHW). Original reference: https://www.nature.com/articles/s41586-020-2534-z

 - citation: Jacox, Michael G., et al. "Thermal displacement by marine heatwaves." Nature 584.7819 (2020): 82-86.

 - Original code ref (in Matlab): https://github.com/mjacox/Thermal_Displacement/

``` {r required_package, warning=FALSE, message=FALSE, results='hide'}
if (!require("curl")) install.packages("curl"); library(curl)
if (!require("magrittr")) install.packages("magrittr"); library(magrittr)
if (!require("data.table")) install.packages("data.table"); library(data.table)

```
First we download dataset from NOAA OISST v2.1. The dataset metadata: https://www.ncei.noaa.gov/products/optimum-interpolation-sst 

 - dataset citation: Huang, B., C. Liu, V. Banzon, E. Freeman, G. Graham, B. Hankins, T. Smith, and H.-M. Zhang, 2020: Improvements of the Daily Optimum Interpolation Sea Surface Temperature (DOISST) Version 2.1, Journal of Climate, 34, 2923-2939. doi: 10.1175/JCLI-D-20-0166.1
 - Note that 19880101 in filename means date of this data file is 1988/01/01
 - "nc" file extension means NetCDF


```{r testfile_of_oisst}
baseurl <- "https://www.ncei.noaa.gov/data/sea-surface-temperature-optimum-interpolation/v2.1/access/avhrr"

# define a file download save path
monthi<- "201109"
datei <- paste0(monthi, "01")
fileprex <- "oisst-avhrr-v02r01"
filename <- paste(fileprex, datei, "nc", sep=".")
oisstfile <- paste(baseurl, monthi, filename, sep="/")  
print(paste0("Download from: ", oisstfile))
# "https://www.ncei.noaa.gov/data/sea-surface-temperature-optimum-interpolation/v2.1/access/avhrr/198801/oisst-avhrr-v02r01.19880101.nc"

dest  <- "../data_src/oisst/daily/"
fileo <- paste0(dest, filename)
print(paste0("File output: ", fileo))

if (!file.exists(fileo)) {
  tryCatch({
    curl::curl_download(oisstfile, destfile = fileo)
  }, error = function(e) paste0(datei, ": ", e))
}  
```

Use R package "stars" to read .nc file (BTW, we used "ncdf4" package in the past)

 - ref: https://r-spatial.github.io/stars/reference/read_stars.html
 
``` {r required_stars, warning=FALSE, message=FALSE}
if (!require("stars")) install.packages("stars"); library(stars)
if (!require("abind")) install.packages("abind"); library(abind)
if (!require("dplyr")) install.packages("dplyr"); library(dplyr)

sts <- read_stars(fileo)
str(sts)
```

``` {r plot_stars_sst, warning=FALSE, message=FALSE, results='hide'}
if (!require("ggplot2")) install.packages("ggplot2"); library(ggplot2)
if (!require("ggExtra")) install.packages("ggExtra"); library(ggExtra)
#if (!require("viridis")) install.packages("viridis"); library(viridis)

var <- "sst" #use abind::adrop to drop undesired dimentions except sst
z <- sts %>% dplyr::select(var) %>% adrop 
df <- as.data.frame(z)
#class(df[,var])
#[1] "units"
df[,var] <- unclass(df[,var])
#class(df[,var])
#[1] "numeric"

# plot settings
# range(na.omit(df[,var]))
# -1.80 31.23
minz <- -2; maxz <- 32.5
legend_direction <- "horizontal"
# a custom color palette used in SST
jet.colors <-
  colorRampPalette(c("#00007F", "blue", "#007FFF", "cyan",
                     "#7FFF7F", "yellow", "#FF7F00", "red", "#7F0000"))

gx <- ggplot() +  
    geom_tile(data=df, aes_string(x="x", y="y", fill=var), alpha=0.8) + 
    xlab(NULL) + ylab(NULL)
# data range and color scale setup
gx <- gx + 
    coord_sf() + #xlim=c(0, 360), ylim=c(-90, 90)) +
    scale_x_continuous(limits = c(0, 360), expand = c(0, 0)) + 
    scale_y_continuous(limits = c(-90, 90), expand = c(0, 0)) +
    scale_fill_gradientn(toupper(var), colors = jet.colors(16), limits=c(minz, maxz),
                       breaks=seq(100*minz, 100*maxz, by = as.integer((maxz-minz)*100/3))/100)

# plot panel setup
gx <- gx + 
    ggExtra::removeGrid(x=TRUE, y=TRUE) +
    theme(
      panel.background = element_blank(), 
      panel.border = element_rect(colour = "black", fill=NA, size=0.75),
      panel.grid.major.x = element_blank(),
      panel.grid.major.y = element_blank(),
      panel.grid.minor = element_blank(),
      plot.margin = unit(c(0,0,0,0), "cm"),
      panel.spacing = unit(0,"cm"),
      strip.background = element_blank(),
      strip.text.y = element_text(angle = 0),
      axis.text.x  = element_text(family = "sans"),
      axis.title.x = element_text(family = "sans"),
      axis.title.y = element_text(family = "sans"),
      axis.text.y = element_text(family = "sans"),
      legend.key = element_blank(),
      legend.key.size = unit(0.8,"line"),
      legend.box.background = element_blank(),
      legend.title = element_text(size=10),
      legend.text =  element_text(family = "sans", size=7),
      legend.background = element_rect(fill = "transparent", colour = "transparent"),
      legend.direction=legend_direction,
      legend.key.height = unit(fifelse(legend_direction=="horizontal",0.2,2),'cm'),
      legend.position = "bottom"
    )
gx

```


``` {r read_sst_anomaly, warning=FALSE, message=FALSE, results='hide'}
var <- "anom" #use abind::adrop to drop undesired dimentions except anom = sst anomaly
za <- sts %>% dplyr::select(var) %>% adrop 
da <- as.data.frame(za)
da[,var] <- unclass(da[,var])
setDT(da)[1:10,]
```

``` {r plot_sst_anomaly, warning=FALSE, message=FALSE, results='hide'}
minz <- -4; maxz <- 6
anom.colors <-
  colorRampPalette(c("#00007F", "blue", "#007FFF", "#00d0ff", "#83e6fc", "white", 
                     "yellow", "#FF8800", "#ff5e00", "#ff3700", "red", "#7F0000"))
ga <- ggplot() +  
    geom_tile(data=da, aes_string(x="x", y="y", fill=var), alpha=0.8) + 
    xlab(NULL) + ylab(NULL)
# data range and color scale setup
ga <- ga + 
    coord_sf() + 
    scale_x_continuous(limits = c(0, 360), expand = c(0, 0)) + 
    scale_y_continuous(limits = c(-90, 90), expand = c(0, 0)) +
    scale_fill_gradientn(toupper(var), colors = anom.colors(16), limits=c(minz, maxz),
                       breaks=c(-4.0, -2.0, -1.5, -1.0, -0.5, 0, 0.5, 1.0, 1.5, 2.0, 4.0, 6.0))

# plot panel setup
ga <- ga + 
    ggExtra::removeGrid(x=TRUE, y=TRUE) +
    theme(
      panel.background = element_blank(), 
      panel.border = element_rect(colour = "black", fill=NA, size=0.75),
      panel.grid.major.x = element_blank(),
      panel.grid.major.y = element_blank(),
      panel.grid.minor = element_blank(),
      plot.margin = unit(c(0,0,0,0), "cm"),
      panel.spacing = unit(0,"cm"),
      strip.background = element_blank(),
      strip.text.y = element_text(angle = 0),
      axis.text.x  = element_text(family = "sans"),
      axis.title.x = element_text(family = "sans"),
      axis.title.y = element_text(family = "sans"),
      axis.text.y = element_text(family = "sans"),
      legend.key = element_blank(),
      legend.key.size = unit(0.6,"line"),
      legend.box.background = element_blank(),
      legend.title = element_text(size=10),
      legend.text =  element_text(family = "sans", size=6),
      legend.background = element_rect(fill = "transparent", colour = "transparent"),
      legend.direction="horizontal",
      legend.key.height = unit(0.2, 'cm'),
      legend.key.width = unit(1.5, 'cm'),
      legend.position = "bottom"
    )
ga

```